---
title: "NCBIDataModeling"
author: "Simon Liles"
date: "4/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading in the data. 
NCBI has a lot of different data that is associated with a single organism. One of the many files that comes from the NCBI data is a report, for example with COVID-19 there is a Virus Report file. The file isin JSON List format which puts an entire JSON file on a single line. Each individual JSON file contains lots of metadata regarding research on the virus, mostly on its genome, and there are several fields that are shared in many of the records, but not all fields are included in every record. 

To read in the Virus Report data, I use the `report_as_dataframe()` function that is included in the protein8k package that I have written. The code would be as follows. 

```{r, results='hide'}
library(protein8k)

#String representing the file path from the working directory to the report
jsonListName <- "data/covid19_data/data_report.jsonl"

#Convert JSONL into a single large data frame
report_df <- report_as_dataframe(fromJSONL(jsonListName))
```

Inside the call to `report_as_dataframe()`, I make a call to `fromJSONL()` which is a function included as a part the protein8k package. All it does is parse each JSON file contained in the JSONL file, and makes it into a list. 

## Extra steps to prepare the data
While `report_as_dataframe()` does most of the preprocessing steps for me, I need to take a couple of extra steps for this more specific use case of the data. 

First I need to remove extraneous information from the geo_Location column. This contains information about which country the data came from in a region, but it also sometimes contain information about which territory within that region the data came from. The territory information will only cause the final plot to be more cluttered, so I remove it here. Based on the formatting rules I know that country names and territory names are separated by a colon, thus I remove everything after that point. 

Isolate source is stored in the original data frame as a factor data which can be useful for other visualizations, but for this one it will need to be character data. 

Last I add a column called root, and it is just filled with the string "root" for all records. This will serve to tie all the branches together in the final visualization. 

```{r}
report_clean_df <- report_df
report_clean_df$geo_Location <- gsub(":.*", "", report_df$geo_Location)
report_clean_df$isolate_source <- as.character(report_clean_df$isolate_source)
report_clean_df$root <- "root"
```

## Making Circle Packing Visualization
With a circle packing visualization it is possible to visualize the hierachal relationships that exist within large data sets. Here I use the ggpraph package to make the visualization. 

First I will load all the libraries I will need for making the visualization
```{r, message=FALSE, warning=FALSE}
#Load libraries
library(ggraph)
library(igraph)
library(tidyverse)
```

### Make the network
To create a circle packing visualization with ggpraph, first I need to make a network object. This will invlove making a list of nodes and an edge list. Then I can put it together into an igraph network object before plotting it as a circle packing visualization. 

#### Make a nodes list
The first step in making a network is creating the nodes list. From a data frame this fairly simple, I first squeeze the selected columns into a single column before counting the frequency of each node. I will use the frequencies later to drive the size of the circles. As a final step, I add another column for unique ids for every node. 

```{r}
#Make list of nodes with frequencies for sizes
nodes <- report_clean_df %>%
  pivot_longer(cols = c(root, isolate_source, geo_Region, geo_Location), 
               values_to = "label") %>%
  count(label, name = "size") %>% 
  rowid_to_column("id")
```

Now I will do a couple of things to make the final visualization a little nicer looking. First I make the root node size 1, otherwise it will just be equal to the number of records. Next I will make the scale log 10. This will make it easier to differentiate differences at the bottom of the scale. 

```{r}
nodes$size[nodes$label == "root"] <- 1

nodes$size <- log10(nodes$size)
```

#### Make the edge list
The edge list is what will make the network, it describes all the connections, or in the case of this visualization, the hierarchical relationships. First I create a list of four element vectors, each element being a value from the cleaned data frame. This is turned into a 4 column matrix before adjacent columns are binded together to make an edge list. I then replace the names with IDs from the nodes list. 

```{r}
#First create a list with an explicit hierarchy
hierarchy_list <- list()

for(record in 1:nrow(report_clean_df)) {
  branch <- c(report_clean_df$root[record],
              report_clean_df$isolate_source[record],
              report_clean_df$geo_Region[record], 
              report_clean_df$geo_Location[record])
  hierarchy_list[[record]] <- branch
}

# create edge list (from in first column / to in the second)
d <- do.call(rbind, hierarchy_list)
edges <- rbind(d[,1:2], d[,2:3], d[,3:4])

edges <- as.data.frame(edges)

edges <- edges %>%
  left_join(nodes, by = c("V1" = "label")) %>%
  rename(from = id)

edges <- edges %>%
  left_join(nodes, by = c("V2" = "label")) %>%
  rename(to = id)
```

The final step is to clean up the edge list so that it only has the IDs. 
```{r}
#Clean up edges dataframe
edges <- select(edges, from, to)
```

#### Make the network object
As a final step, I combine all the previous work to make a network object. This will be plotted in the next step. 

```{r}
###Making the network ##########################################################
loc_graph <- graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)
```

### Making the visual
The code to make the visualization looks complicated, however it actually quite simple. First I make the call to ggraph, specify circle packing layout and give the weights for the size of the nodes. Then I color it based on depth inside the hierarchy, and I add labels for the circles. 

```{r}
ggraph(loc_graph, 'partition', circular = TRUE) + 
  geom_node_arc_bar(aes(fill = size, color = depth), size = 0.25) +
  geom_node_text(aes(label = label), size = 2.5) + 
  theme_void() +
  theme(legend.position = "right") + 
  scale_color_continuous(guide = "none") +
  scale_fill_viridis_c(direction = 1)
```

