---
title: "All Protein8k Modeling"
author: "Simon Liles"
date: "5/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Working with Protein Data Bank Files

## Loading a PDB file
```{r}
library(protein8k)

fileName <- "data/1aieH"

my_protien <- read.pdb(fileName)
```

## Plotting a Protein Object
PDB files include a lot of data on a protein, however tables of data are not very informative. With the protein8k package it is possible to create several different visualizations of Protein Objects. 

### Generating 3D Plots
To get an understanding of the structure of the protein, it is possible to make a 3D plot using the atomic record. 

```{r}
#Generates a static 3D plot of a protein
plot3D(my_protien)
```

This plot can then be animated by setting the `animated` argument to `TRUE`. 

```{r}
#Animate the protein structure spinning
plot3D(my_protien, animated = TRUE, type = "p", groups = residue_name, 
       image_width = 300, image_height = 300)
```

### Generating smoothing models of Protein Structure
Another useful visualization is to apply smoothing lines to the point in the atomic record. With the `plotModels()` function, one can create a series of visualizations that look at each plane of the structure and apply a smoothing line. 

```{r message=FALSE,error=FALSE,warning=FALSE}
#Generate a model of each plane of the protein structure
plotModels(my_protien)
```

# Working with NCBI Virus Report Data

## Loading in the data. 
NCBI has a lot of different data that is associated with a single organism. One of the many files that comes from the NCBI data is a report, for example with COVID-19 there is a Virus Report file. The file isin JSON List format which puts an entire JSON file on a single line. Each individual JSON file contains lots of metadata regarding research on the virus, mostly on its genome, and there are several fields that are shared in many of the records, but not all fields are included in every record. 

To read in the Virus Report data, I use the `report_as_dataframe()` function that is included in the protein8k package that I have written. The code would be as follows. 

```{r, results='hide'}
library(protein8k)

#String representing the file path from the working directory to the report
jsonListName <- "data/covid19_data/data_report.jsonl"

#Convert JSONL into a single large data frame
report_df <- report_as_dataframe(fromJSONL(jsonListName))
```

Inside the call to `report_as_dataframe()`, I make a call to `fromJSONL()` which is a function included as a part the protein8k package. All it does is parse each JSON file contained in the JSONL file, and makes it into a list. 

## Extra steps to prepare the data
While `report_as_dataframe()` does most of the preprocessing steps for me, I need to take a couple of extra steps for this more specific use case of the data. 

First I need to remove extraneous information from the geo_Location column. This contains information about which country the data came from in a region, but it also sometimes contain information about which territory within that region the data came from. The territory information will only cause the final plot to be more cluttered, so I remove it here. Based on the formatting rules I know that country names and territory names are separated by a colon, thus I remove everything after that point. 

Isolate source is stored in the original data frame as a factor data which can be useful for other visualizations, but for this one it will need to be character data. 

Last I add a column called root, and it is just filled with the string "root" for all records. This will serve to tie all the branches together in the final visualization. 

```{r}
report_clean_df <- report_df
report_clean_df$geo_Location <- gsub(":.*", "", report_df$geo_Location)
report_clean_df$isolate_source <- as.character(report_clean_df$isolate_source)
report_clean_df$root <- "root"
```

## Making Circle Packing Visualization
With a circle packing visualization it is possible to visualize the hierachal relationships that exist within large data sets. Here I use the ggpraph package to make the visualization. 

First I will load all the libraries I will need for making the visualization
```{r, message=FALSE, warning=FALSE}
#Load libraries
library(ggraph)
library(igraph)
library(tidyverse)
```

### Make the network
To create a circle packing visualization with ggpraph, first I need to make a network object. This will invlove making a list of nodes and an edge list. Then I can put it together into an igraph network object before plotting it as a circle packing visualization. 

#### Make a nodes list
The first step in making a network is creating the nodes list. From a data frame this fairly simple, I first squeeze the selected columns into a single column before counting the frequency of each node. I will use the frequencies later to drive the size of the circles. As a final step, I add another column for unique ids for every node. 

```{r}
#Make list of nodes with frequencies for sizes
nodes <- report_clean_df %>%
  pivot_longer(cols = c(root, isolate_source, geo_Region, geo_Location), 
               values_to = "label") %>%
  count(label, name = "size") %>% 
  rowid_to_column("id")
```

Now I will do a couple of things to make the final visualization a little nicer looking. First I make the root node size 1, otherwise it will just be equal to the number of records. Next I will make the scale log 10. This will make it easier to differentiate differences at the bottom of the scale. 

```{r}
nodes$size[nodes$label == "root"] <- 1

nodes$size <- log10(nodes$size)
```

#### Make the edge list
The edge list is what will make the network, it describes all the connections, or in the case of this visualization, the hierarchical relationships. First I create a list of four element vectors, each element being a value from the cleaned data frame. This is turned into a 4 column matrix before adjacent columns are binded together to make an edge list. I then replace the names with IDs from the nodes list. 

```{r}
#First create a list with an explicit hierarchy
hierarchy_list <- list()

for(record in 1:nrow(report_clean_df)) {
  branch <- c(report_clean_df$root[record],
              report_clean_df$isolate_source[record],
              report_clean_df$geo_Region[record], 
              report_clean_df$geo_Location[record])
  hierarchy_list[[record]] <- branch
}

# create edge list (from in first column / to in the second)
d <- do.call(rbind, hierarchy_list)
edges <- rbind(d[,1:2], d[,2:3], d[,3:4])

edges <- as.data.frame(edges)

edges <- edges %>%
  left_join(nodes, by = c("V1" = "label")) %>%
  rename(from = id)

edges <- edges %>%
  left_join(nodes, by = c("V2" = "label")) %>%
  rename(to = id)
```

The final step is to clean up the edge list so that it only has the IDs. 
```{r}
#Clean up edges dataframe
edges <- select(edges, from, to)
```

#### Make the network object
As a final step, I combine all the previous work to make a network object. This will be plotted in the next step. 

```{r}
###Making the network ##########################################################
loc_graph <- graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)
```

### Making the visual
The code to make the visualization looks complicated, however it actually quite simple. First I make the call to ggraph, specify circle packing layout and give the weights for the size of the nodes. Then I color it based on depth inside the hierarchy, and I add labels for the circles. 

```{r}
ggraph(loc_graph, 'partition', circular = TRUE) + 
  geom_node_arc_bar(aes(fill = size, color = depth), size = 0.25) +
  geom_node_text(aes(label = label), size = 2.5) + 
  theme_void() +
  theme(legend.position = "right") + 
  scale_color_continuous(guide = "none") +
  scale_fill_viridis_c(direction = 1)
```

## Using the data to make Alluvial Plots
Sometimes an Alluvial Plot may be better than a Circular Partition Plot for visualizing the relationships between different layers of a hierarchy. 

### Preparing the Data
For the alluvial plot I will want my data to be in 3 columns, each column being a layer in the hierarchy. Then I will add a fourth column for the frequency of each unique row. 

Here is the library I will be using to build the plot. This library is built to integrate with ggplot which will make the plotting easier. 

```{r}
#Making an alluvial plot
library(ggalluvial)
```

To begin I will use the matrix d that I generated before for the circular plot. For this plot I do not want the root node, So I will remove it now before converting the matrix to a data frame.

```{r}
#Prepare a new edge list for the alluvial plot
edge_alvl <- d
edge_alvl <- edge_alvl[,2:4]
edge_alvl <- as.data.frame(edge_alvl)
```

The next step then is to use the dplyr library to add the weights or frequency column, and then rename the columns from their default names to names that are more meaningful. 

```{r}
#Add weights column and rename columns to be more meaningful
edge_alvl <- group_by(edge_alvl, V1, V2, V3) %>%
  count() %>%
  ungroup() %>%
  rename(source = V1) %>%
  rename(region = V2) %>%
  rename(location = V3) %>%
  rename(weight = n)
```

Now that the data frame is mostly set, I can begin working on cleaning it so that the final plot can be more meaningful. First I remove incomplete cases, or records with null values, in the source column. In the final plot these represented a large portion of sources and distracted from the data that held more meaning. The enxt thing that needs to be changed is the scale for the weights. Many of the values are close to 0, while there are a few that are over a thousand, to correct for this in the visualization, I adjjust it to the log 10 scale. 

```{r}
#Remove NA values from source column
edge_alvl <- edge_alvl[complete.cases(edge_alvl$source),]

#Make weights log10 scale
edge_alvl$weight <- log10(edge_alvl$weight)
```

Before creating the visualization, it is important to check that you have the proper form. If this line returns `TRUE`, then you have the proper form for the data. 

```{r}
#Check the form
is_alluvia_form(edge_alvl)
```

Now I can create the plots using the following code. I create two plots, each colors the flows slightly differetnly based on which column I set for it to be colored by. 

```{r}
#Plot with ggalluvial
#Color by source
ggplot(edge_alvl, 
       aes(y = weight, axis1 = source, axis2 = region, axis3 = location)) +
  geom_alluvium(aes(fill = source), width = 0, knot.pos = 0, reverse = FALSE) +
  guides(fill = FALSE) + 
  geom_stratum(width = 1/8, reverse = FALSE) + 
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE, 
            size = 2) 

#Color by region
ggplot(edge_alvl, 
       aes(y = weight, axis1 = source, axis2 = region, axis3 = location)) +
  geom_alluvium(aes(fill = region), width = 0, knot.pos = 0, reverse = FALSE) +
  guides(fill = FALSE) + 
  geom_stratum(width = 1/8, reverse = FALSE) + 
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE, 
            size = 2) 
```